### 2016-01-27 doc/manuscript.Rmd
Started manuscript by adding methods, intro outline, and some results

---
bibliography: 
    - /home/luke/Documents/bibliography/bib/bib-db/myrefs.bib
    - /home/luke/Documents/bibliography/bib/bib-db/nefaDiabetes.bib
csl:
    - /home/luke/Documents/bibliography/styles/american-medical-association-no-et-al.csl
output:
    word_document:
      reference_docx: /home/luke/Templates/journal-templates/ajcn-template.docx
      fig_height: 6 
      fig_width: 5 
      fig_caption: true
---

@@

@@

@@

<!--
## Load data ##
-->

```{r setup, message = FALSE, echo = FALSE}
invisible(sapply(list.files('R', pattern = '*.R', full.names = TRUE), source))
run_setup()
load_data()
dim(ds)
```

```{r caption_setup}
## Include captions below using `captioner` package
figNums <- captioner(prefix = 'Figure')
supFigNums <- captioner(prefix = 'Supplementary Figure')
tabNums <- captioner(prefix = 'Table')
supTabNums <- captioner(prefix = 'Supplementary Table')
```

@@

After running the GEE models for the longitudinal analysis, linear discriminant
analysis (LDA) {{or QDA}} was used to identify the patterns of NEFA composition for
tertiles of the metabolic disorders. Jacknifed cross-validation method used in
the LDA analysis. Only the baseline visit data was used for the LDA analysis and
only the outcome and predictor variables could be included in the analysis.

@@

# Results #

# Discussion #

@@

```{r figDist}
ds %>%
    filter(VN == 0) %>%
    select(SID, matches('^ne\\d')) %>%
    filter(complete.cases(.)) %>% 
    gather(Measure, Value, -SID) %>% 
    mutate(Measure = renaming_fa(Measure) %>% 
               gsub('ne_', '', .) %>%
               factor(., levels = unique(.))) %>% 
    spread(Measure, Value) %>% 
    select(-SID) %>% 
    trance('boxes_dots') %>% 
    visualize(dots = FALSE,
              xlab = 'Concentration (nmol/mL)',
              ylab = 'Non-esterified fatty acid') %>% 
    vision_simple(base_size = 10)
```

```{r geeData}

outcomes <- c('linvHOMA', 'lISI', 'lIGIIR', 'lISSI2')
ne_pct <- grep('^pct_ne', names(ds), value = TRUE)
ne_conc <- c('ne_Total', grep('^ne\\d\\d', names(ds), value = TRUE))
covariates <- c('Waist', 'VN', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'FamHistDiab')

gee_df <- full_join(
    ds %>%
        select(-matches('pct_ne|^ne'),-ne_Total),
    ## Scale all the fatty acids
    ds %>%
        filter(VN == 0) %>%
        select(SID, ne_Total, matches('pct_ne|^ne')) %>%
        mutate_each(funs(as.numeric(scale(.))),-SID),
    by = 'SID') %>%
    group_by(VN) %>%
    mutate(
        Waist = as.numeric(scale(Waist)),
        ALT = as.numeric(scale(ALT)),
        BaseAge = as.numeric(scale(BaseAge))
        ) %>% 
    ungroup() %>%
    arrange(SID, VN)

gee_results <- bind_rows(
    analyze_gee(gee_df, outcomes, ne_pct, covariates, 'mol%'),
    analyze_gee(gee_df, outcomes, ne_conc, covariates, 'nmol/mL')
    ) %>% 
    mutate()

gee_results_unadj <- bind_rows(
    analyze_gee(gee_df, outcomes, ne_pct, covariates, 'mol%'),
    analyze_gee(gee_df, outcomes, ne_conc, covariates, 'nmol/mL')
    )

@@

```{r figGee}
gee_results %>%
    trance('main_effect') %>%
    visualize(groups = 'unit~Yterms',
              xlab = 'Percent difference with 95% CI in the outcomes\nfor each SD increase in fatty acid',
              ylab = 'Non-esterified fatty acids') %>% 
    vision_simple(legend_position = 'bottom') +
    theme(
        axis.text.y = element_text(face = ifelse(
        gee_results$Xterms == 'Total', "bold", "plain"
        )),
        legend.key.width = grid::unit(0.75, "line"),
        legend.key.height = grid::unit(0.75, "line"),
        panel.margin = grid::unit(0.75, "lines")
        ) +
    scale_alpha_discrete(name = 'P-value', range = c(0.4, 1.0))
```

```{r supTabQIC}

subds <- ds %>%
  select(SID, VN, lISSI2, pl_Total, Sex, Ethnicity, MET, ALT,
         BaseAge, Waist, BMI, AlcoholPerWk, TobaccoUse, SelfEdu,
         Occupation, totalNE, FamHistDiab) %>%
  mutate(SelfEdu = as.factor(SelfEdu),
         Occupation = as.factor(Occupation),
         TobaccoUse = as.factor(TobaccoUse),
         AlcoholPerWk = as.factor(AlcoholPerWk)) %>%
  filter(complete.cases(.))

levels(subds$SelfEdu)[1:2] <- c('2', '2')
subds$SelfEdu <- droplevels(subds$SelfEdu)

M0 <- geeglm(lISSI2 ~ pl_Total + VN, data = subds,
                id = SID, family = gaussian,
                corstr = 'exchangeable')
M1 <- update(M0, . ~ . + Sex + Ethnicity + BaseAge)
M2 <- update(M0, . ~ . + Sex + Ethnicity + BaseAge + Waist)
M3 <- update(M0, . ~ . + Sex + Ethnicity + BaseAge + BMI)
M4 <- update(M0, . ~ . + Sex + Ethnicity + BaseAge + Waist + MET +
               AlcoholPerWk + TobaccoUse + SelfEdu + Occupation +
               FamHistDiab)
M5 <- update(M0, . ~ . + Sex + Ethnicity + BaseAge + MET + Waist +
               AlcoholPerWk + TobaccoUse + SelfEdu + Occupation +
               FamHistDiab + totalNE + ALT)
M6 <- update(M0, . ~ . + Sex + Ethnicity + Waist + totalNE)
M7 <- update(M0, . ~ . + Sex + Ethnicity + Waist + totalNE + ALT)
M8 <- update(M0, . ~ . + Sex + Ethnicity + Waist + totalNE + ALT + BaseAge)
M9 <- update(M0, . ~ . + Sex + Ethnicity + Waist + totalNE + ALT + BaseAge + FamHistDiab)

set.caption(supTabRef('tabQIC', 'Comparing generalized estimating equation
models adjusting for different covariates using
Quasi-Likelihood Information Criterion.'))
MuMIn::model.sel(M0, M1, M2, M3, M4, M5, M6, M7, M8, M9,
          rank = QIC) %>%
  tableModelSelection()

tableModelSelection <- function(selection.data) {
    selection.data %>%
      add_rownames() %>%
      select(Model = rowname, QIC, Delta = delta) %>%
      pander(digits = 2)
}

>>>>

---
bibliography: 
    - /home/luke/Documents/bibliography/bib/bib-db/nefaDiabetes.bib
    - /home/luke/Documents/bibliography/bib/bib-db/myrefs.bib
csl:
    - /home/luke/Documents/bibliography/styles/american-medical-association-no-et-al.csl
output:
    word_document:
      reference_docx: /home/luke/Templates/journal-templates/ajcn-template.docx
      fig_height: 6 
      fig_width: 5 
      fig_caption: true
---

@@

Popular hypothesis: obese -> FFA -> IR.. may not be the case

@@

Association between higher FFA and muscle/hepatic IR, but challenging as adipose
tissue can also be IR. So is it the higher FFA -> IR or is it IR -> FFA?

@@

```{r setup, collapse=TRUE}
knitr::opts_knit$set(root.dir = '../')
knitr::opts_chunk$set(echo = FALSE)
```

```{r loadData, message=FALSE}
source('.Rprofile')
suppressMessages(run_setup())
load_data()
dim(ds)
```

```{r caption_setup}
## Include captions below using `captioner` package
fig <- captioner(prefix = 'Figure')
cite_f <- pryr::partial(fig, display = 'cite')
sfig <- captioner(prefix = 'Supplementary Figure')
cite_sf <- pryr::partial(sfig, display = 'cite')
tab <- captioner(prefix = 'Table')
cite_t <- pryr::partial(tab, display = 'cite')
stab <- captioner(prefix = 'Supplementary Table')
cite_st <- pryr::partial(stab, display = 'cite')
# usage: cite_st('basicChar')

tab_basic <- tab('basicChar', 'Basic characteristics of the PROMISE participants at each of the 3 time points.')
fig_dist <- fig('nefaDist', 'Concentrations (nmol/mL) of each non-esterified fatty acid in PROMISE participants at the baseline visit (2004-2006)')
fig_gee <- fig('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function using generalized estimating equations over the 6 years in the PROMISE cohort.  Adjusted for {{covariates}}. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.  P-values presented were adjusted for the false discovery rate.')
```

@@

Sensitivity analysis adjusting for clinically measured TG, as TG at the 6 year
visit was only done on a subset of participants.

While the GEE modeling takes into consideration the longitudinal design of the
data, a limitation is that it can not analyze the inherent multivariate nature
of the composition of the NEFA fraction. As such, after running the GEE models
for the longitudinal analysis, linear discriminant analysis (LDA) {{or QDA}} was
used to identify the patterns of NEFA composition for tertiles of the metabolic
disorders. LDA is not suited for longitudinal data and so only the baseline
visit data was used and only the outcome and predictor variables could be
included in the analysis. Data from the later visit dates were used to confirm
the results of the LDA analysis.

@@

# Results

```{r geeData, cache=TRUE}

outcomes_is <- c('linvHOMA', 'lISI')
outcomes_bcf <- c('lIGIIR', 'lISSI2')
outcomes <- c(outcomes_is, outcomes_bcf)
ne_pct <- grep('^pct_ne', names(ds), value = TRUE)
ne_conc <- c('ne_Total', grep('^ne\\d\\d', names(ds), value = TRUE))
covariates_is <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT')
covariates_bcf <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 
                    'MET', 'AlcoholPerWk')

gee_df <- get_gee_data(ds)
gee_results <- bind_rows(
    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is, 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is, 'nmol/mL', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf, 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf, 'nmol/mL', adj.p = TRUE)
    )

gee_results_unadj <- bind_rows(
    analyze_gee(gee_df, outcomes, ne_pct, 'VN', 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes, ne_conc, 'VN', 'nmol/mL', adj.p = TRUE)
    ) %>% 
    mutate(Yterms = factor(Yterms, unique(Yterms)))
```

```{r ldaData}
# include also the 3-yr and 6-yr data for training/testing.
lda_df <- ds %>%
    filter(VN == 0) %>%
    mutate(BCF = rstatsToolkit::aide.tertile(lISSI2) %>%
               factor(., labels = c('1st tertile: log(ISSI-2)', 
                                    '2nd tertile: log(ISSI-2)', 
                                    '3rd tertile: log(ISSI-2)'), 
                      ordered = TRUE)) %>%
    select(BCF, matches('^ne\\d\\d')) %>%
    mutate_each(funs(as.numeric(scale(.))), -BCF) %>%
    na.omit() %>% 
    arrange(BCF)
lda_results <- lda(BCF ~ ., data = lda_df)
```

```{r resultsInline, dependson='geeData'}
# extract gee beta coefficient for indicating magnitude
pctChg.o <- ds %>%
  select(f.VN, HOMA, ISI, IGIIR, ISSI2) %>%
  gather(Measure, Value, -f.VN) %>%
  na.omit() %>%
  group_by(Measure, f.VN) %>%
  summarise(med = median(Value)) %>%
  spread(f.VN, med) %>%
  mutate(pctChg = ((`6-yrs` - `0-yr`) / `0-yr`) * 100) %>%
  select(pctChg) %>%
  abs() %>%
  round(1) %>%
  {paste0(min(.), '% to ', max(.), '%')}

cumPctPL <- calcCumPctFA(ds, 'pct_pl', 90)
calcCumPctFA <- function(data, variable.pattern, filter.value) {
    data %>%
      filter(VN == 0) %>%
      select(matches(variable.pattern)) %>%
      gather(fat, value) %>%
      group_by(fat) %>%
      summarise(pct = mean(value, na.rm = TRUE)) %>%
      arrange(desc(pct)) %>%
      mutate(cumpct = cumsum(pct) %>% round(1) %>% format(nsmall = 1),
             pct = round(pct, 1) %>% format(nsmall = 1)) %>%
      filter(cumpct <= filter.value)
}
extractBeta <- function(data) {
    data %>%
      mutate(estCI = paste0('(beta: ', trim(format(round(estimate, 1), nsmall = 1)),
                            ' CI: ',
                            trim(format(round(conf.low, 1), nsmall = 1)),
                            ', ',
                            trim(format(round(conf.high, 1), nsmall = 1)),
                            ')'),
             #indep.estCI = paste(indep, estCI),
             #dep = paste(dep %>% gsub('log\\((.*)\\)', '\\1', .), estCI),
             dep = paste(dep %>% gsub('log\\((.*)\\)', '\\1', .))
             ) %>%
      filter(p.value <= 0.05) %>%
      arrange(dep, desc(estimate)) %>%
      select(dep, indep, estCI)
}
```

## Basic characteristics of the PROMISE cohort

## Distribution of NEFA

## Correlations of NEFA with basic characteristics

## GEE

As shown in {{supFigGeeUnadj}}, several associations were seen in the unadjusted
GEE models, particularly for the concentration data. Full model adjustment
attenuated most of these associations, with total NEFA, 16:0, and 18:2n-6 (all
as nmol/mL) had negative associations between IGI/IR and ISSI-2 ({{figGee}}).
The magnitude of association for each of these variables was fairly consistent
for each beta-cell function measure. For every one SD decrease in any of these
three NEFA variables, there was a predicted {{magnitude}}% lower IGI/IR and a
{{magnitude}}% lower ISSI-2 at each clinic visit. None of the NEFA variables as
a mol% had any association with the outcomes.

Adjusting for TAG attenuates all associations with IGI/IR and ISSI-2 (data not
shown). This may be due to either or both of two factors, mediation of TAG on
the pathway between NEFA and beta-cell function or the lower sample size at the
6 year visit from the incomplete measurement of TAG due to cuts in funding.

There were no significant interaction effects between time and the individual
NEFA on any of the outcome measures (data not shown).

## Discriminant analysis

# Discussion

Brief overall conclusion...

Given we found associations with the nmol/mL data, but none in the mol% data,
this suggests to us that the total size of the NEFA fraction is more important
in the pathogenesis of diabetes than is the specific FA composition. A
limitation of the mol% data is the inherent dependency between the FA as an
increase in one FA by definition means other FA must decrease. In the context of
this dependency, the associations we found for 16:0 and 18:2n-6 as nmol/mL
suggests that these NEFA may be contributing to the harmful association seen for
the total NEFA fraction, irrespective of changes in the other NEFA.

We found that higher palmitic acid (16:0) and linoleic acid (18:2n-6) associated
with lower beta-cell function by {{number?}}. Expand on experimental studies,
pathology/biology, mechanisms, etc.

There is substantial experimental research on the role of increasing NEFA and
increasing IR {{cite}}. Two major limitation of these studies are that they tend
to be fairly short term ({{length of time}}) and they tend to increase NEFA to
supraphysiological concentrations {{confirm}}. Epidemiological research has also
shown a link between higher NEFA and higher IR {{cite}}. However, most of these
studies are cross-sectional in nature and can not address causal direction.
Adipose tissue can become insulin resistant and thus hormone-sensitive lipase
may not respond to the inhibitory action of insulin {{cite}}. As such, IR may
temporally occur sooner than elevations in blood NEFA. In this longitudinal
analysis, we saw no association between NEFA composition and neither hepatic IR
(HOMA-IR) nor whole-body insulin sensitivity (ISI). Possible explanations for
this could be that over several years NEFA does not influence hepatic or muscle
IR or that it is IR that elevates NEFA and as such NEFA would not predict IR
over time.

In the sensitivity analysis adjusting for TAG, all associations with beta-cell
function were attenuated. While there is strong experimental evidence on the
role of chronically elevated NEFA, particularly palmitic acid, on impaired
beta-cell function, within a free-living population NEFA may be mediating it's
association through TAG. In normal metabolism, NEFA enters the liver and assists
in the production of TAG that is to be processed into very-low density
lipoproteins (VLDL) {{cite}}. Elevated NEFA may be mediating it's association
through hypertriglyceridemia, a known risk factor for diabetes {{cite}}. A major
caveat with the TAG-adjusted GEE models is the reduced sample size in the 6 year
visit, as only a subset (n={{sample}}) of participants had TAG measured.

Comment on LDA analysis... ...LDA backs up some of the GEE model results,
showing weak discriminatory ability of the NEFA pattern on identifying beta-cell
function... ...While the overall discriminatory ability of the NEFA pattern was
poor, there were certain NEFA that loaded differently on each level of beta-cell
function....

Limitations, etc...

Final, brief conclusion...

@@

```{r figDist}
ds %>%
    filter(VN == 0) %>%
    select(SID, matches('^ne\\d')) %>%
    filter(complete.cases(.)) %>% 
    gather(Measure, Value, -SID) %>% 
    mutate(Measure = renaming_fa(Measure) %>% 
               gsub('ne_', '', .) %>%
               factor(., levels = unique(.))) %>% 
    spread(Measure, Value) %>% 
    select(-SID) %>% 
    trance('boxes_dots') %>% 
    visualize(dots = FALSE,
              xlab = 'Concentration (nmol/mL)',
              ylab = 'Non-esterified fatty acid') %>% 
    vision_simple() +
    theme(axis.ticks.y = element_blank())
```

```{r figGEE, dependson='geeData'}
forest_plot(gee_results)
```

```{r figLDA}
visualize(lda_results, 'means', label.rename.fun = renaming_fa) %>% 
    vision_simple() +
    labs(y = 'Non-esterified fatty acids',
         x = 'Standardized mean of the fatty acid (SD of the nmol/mL)')
```

```{r supFigGeeUnadj, dependson='geeData'}
forest_plot(gee_results_unadj)
```

```{r supTabQIC, eval=FALSE}

@@

```{r supTabGEE}
